Towards a Mechanized Proof of Noninterference

### 2013-12-19 paper/paper.tex
expand if proof in lemma2

In this paper we discuss our experience trying to mechanize a proof of
noninterference for a simple language with security labels, \langName, based on
Pottier and Simonet's \origLang.  This
marks the first step towards writing a verified compiler for a language with
information flow types.  Although we did not complete the full noninterference
proof, we were able to find subtle inaccuracies in the \origLang proofs.  
We discuss what would be involved in completing the proof and how this 
relates to other work.

@@

The last decade has also seen a surge of research in languages which enforce information flow security, such as Jif~\cite{myers1999jflow} or FlowCaml~\cite{simonet2003flow}.  These languages provide features which programmers can use to enforce integrity and confidentiality of data as it is used throughout computations.

@@

The language's type system includes subtyping, which allows data that have a
lower label to flow into higher labelled data.  We use the notation $x \leq y$
to denote that $x$ is a subtype of $y$, whether both $x$ and $y$ are labels or
types (as long as both are the same kind of value).  The rules for subtyping are
given in Figure~\ref{fig:subtype_rules}.  Rule \textsc{S-Int} states that
integers are subtypes of integers with equivalent or higher labels in the label
lattice.  Similarly, rule \textsc{S-Abs} gives the standard subtyping
relationship for functions along with the same label relation.

@@

\textsc{V-Bracket} and \textsc{E-Bracket} both require that there exists a label
from the set $H$ which represents an upward closed sublattice of the labels in
the system representing what is considered to be ``high'' security for the
purposes of the proof of noninterference, discussed later.  These rules require
that the types of each branch of a bracketed value or expression can be typed
with a high security label.  Furthermore, \textsc{E-Bracket} requires that the
$pc$ of the subexpressions must be at least as high as this same label,
requiring that we treat everything in the subexpression as affected by high
security information.

@@

  As a result, we know that $e$ reduces to an answer, because otherwise it would
  be stuck and this would mean that one of the projections is also stuck by
  Lemma~\ref{lem:stuck_bracket}, a contradiction.  Thus, we know that if both
  projections step to an answer, so must the original expression.
\end{proof}

@@

  \begin{case}[lift-if]
    $e = \text{if } <v_1 | v_2> \: e_1 \: e_2$ and $e' = e_i$ for some $i \in
    \{1, 2\}$ in this case.  Applying Lemma~\ref{lem:guard} to the first premise
    of \textsc{E-If} gives $l' \in H$ and $l' \guards t$.  Applying
    Lemma~\ref{lem:projection} to the initial premises of \textsc{E-If} and using
    the result and applying \textsc{E-If} gives $\typeRule{pc \sqcup l}{e_i}$.
    We can then apply \textsc{E-Bracket} to get the goal.
  \end{case}

@@

The noninterference theorem shows that if we have an expression that is typed
with a low label, then replacing any bracketed value in the expression with
either projection will result in the same value for the overall expression.  In
other words, if we vary any high security item in a low security expression, the
value must not be affected.

@@

We found that the process of mechanizing the lemmas that we proved was a useful
exercise in checking our understanding of our language, statements, and the work
of Simonet and Pottier which we based our approach off of.  We found ourselves
often surprised by the number of items that were elided by the proofs in the
paper.  For example, the proof of Lemma~\ref{lem:soundness} is one line of
natural language but turned out to be 40 lines of Coq code, not including all of
the supporting definitions and rules.

\subsection{Inaccuracies in the Original Proof}

We discovered a few small inaccuracies in \cite{InfoFlowML} by running into
issues when formalizing the proofs in Coq.  Listed below are the primary
corrections we were able to make.

\begin{description}
  \item[Non-nesting Brackets in Syntax]
    \hspace{0pt}

    As seen in Figure~\ref{fig:syntax}, we have a slightly different syntax
    structure for \langName from \origLang.  Specifically, we constructed
    values, answers, and expressions explicitly to not allow nesting of
    bracketed values.  This was not how the syntax figures in Simonet and
    Pottier's paper defined this, and as a result, we had trouble when our
    proofs depended on this explicit non-nesting property that was mentioned in
    the prose of their paper.  Realizing that our definitions did not disallow
    this by construction after running into issues in the proof, we were able to
    fix the definition producing what is in Figure~\ref{fig:syntax}.

  \item[Correction of the Soundness Lemma]
    \hspace{0pt}

    Proving our own version of Lemma~\ref{lem:soundness} actually turned out to
    be instructive because it forced us to recognize that the original
    formulation in Simonet and Pottier's paper on \origLang wasn't exactly
    correct as stated.  Specifically, it is entirely possible that a step in the
    operational semantics does not translate to a step in the operational
    semantics under either projection, as in the case of the \ref{eq:lift-app}
    rule, where $\lift{e}_i$ and $\lift{e'}_i$ are identical.  Luckily, we are
    fairly confident, given that we were still able to prove
    Lemma~\ref{lem:stuck_bracket} despite this change, that this is not a
    critical error in the original paper.
\end{description}

\todo{more?}

@@

Once a full Coq proof of noninterference exists, a next step would be to
integrate this proof and all of the related definitions into a verified type
checking program, which could be integrated into a verified compiler.  We
already know that a full verified compiler has been done for C, and this gives
us hope that it would be possible to implement something similar for a language
like Jif.

>>>>

In this paper we discuss our experience trying to mechanize a proof of
noninterference for a simple language with security labels, \langName, based on Pottier and Sionet's \origLang.  This
marks the first step towards writing a verified compiler for a language with
information flow types.  Although we did not complete the full noninterference
proof, we were able to find subtle inaccuracies in the \origLang proofs.  
We discuss what would be involved in completing the proof and how this 
relates to other work.

@@

The last decade has also seen a surge of research in languages which inforce information flow security, such as Jif~\cite{myers1999jflow} or FlowCaml~\cite{simonet2003flow}.  These languages provide features which programmers can use to enforce integrity and confidentiality of data as it is used throughout computations.

@@

The language's type system includes subtyping, which allows data that have a
lower label to flow into higher labelled data.  We use the notatoin $x \leq y$
to denote that $x$ is a subtype of $y$, whether both $x$ and $y$ are labels or
types (as long as both are the same kind of value).  The rules for subtyping are
given in Figure~\ref{fig:subtype_rules}.  Rule \textsc{S-Int} states that
integers are subtypes of integers with equivalent or higher labels in the label
lattice.  Similarly, rule \textsc{S-Abs} gives the standard subtyping
relationship for functions along with the same label relation.

@@

\textsc{V-Bracket} and \textsc{E-Bracket} both require that there exists a label
from the set $H$ which representes an upward closed sublattice of the labels in
the system representing what is considered to be ``high'' security for the
purposes of the proof of noninterference, discussed later.  These rules require
that the types of each branch of a bracketed value or expression can be typed
with a high security label.  Furthermore, \textsc{E-Bracket} requires that the
$pc$ of the subexpressions must be at least as high as this same label,
requiring that we treat everything in the subexpression as affected by high
security information.

@@

  As a result, we know that $e$ reduces to an answer, because otherwise it would
  be stuck and this would mean that one of the projections is also stuck by
  Lemma~\ref{lem:stuck_bracket}, a contradicton.  Thus, we know that if both
  projections step to an answer, so must the original expression.
\end{proof}

@@

  \begin{case}[lift-if]
    $e = \text{if } <v_1 | v_2> \: e_1 \: e_2$ and $e' = e_i$ for some $i \in
    \{1, 2\}$ in this case.  Applying Lemma~\ref{lem:guard} to the first premise
    of \textsc{E-If} gives $l' \in H$ and $l' \guards t$.  Applying
    Lemma~\ref{lem:projection} to the intial premises of \textsc{E-If} and using
    the result and applying \textsc{E-If} gives $\typeRule{pc \sqcup l}{e_i}$.
    We can then apply \textsc{E-Bracket} to get the goal.
  \end{case}

@@

The noninteference theorem shows that if we have an expression that is typed
with a low label, then replacing any bracketed value in the expression with
either projection will result in the same value for the overall expression.  In
other words, if we vary any high security item in a low security expression, the
value must not be affected.

@@

We found that the process of mechanizing the lemmas that we proved was a useful
exercise in checking our understanding of our language, statements, and the work
of Simonet and Pottier which we based our approach off of.  We found ourselves
often surprised by the number of items that were elided by the proofs in the
paper.  For example, the proof of Lemma~\ref{lem:soundness} is one line of
natural languae but turned out to be 40 lines of Coq code, not including all of
the supporting definitions and rules.

Our proof of Lemma~\ref{lem:soundness} actually turned out to be instructive
because it forced us to recognize that the original formulation in Simonet and
Pottier's paper on \origLang~\cite{InfoFlowML} wasn't quite correct.
Specifically, it is entirely possible that a step in the operational semantics
does not translate to a step in the operational semantics under either
projection, as in the case of the \ref{eq:lift-app} rule, where $\lift{e}_i$ and
$\lift{e'}_i$ are identical.  Luckily, we are fairly confident, given that we
were still able to prove Lemma~\ref{lem:stuck_bracket} despite this change, that
this is not a critical error in the original paper.

\todo{More.}

@@

Once a full Coq proof of noninterference exists, a next step would be to
integrate this proof and all of the related definitions into a verified type
checking program, which could be intergrated into a verified compiler.  We
already know that a full verified compiler has been done for C, and this gives
us hope that it would be possible to implement something similar for a language
like Jif.
